\section{Full Parameter Set}\label{app:params}
Table~\ref{tab:fullparams} lists every parameter, complementing the abbreviated
Table~\ref{tab:sensors}. The temperature curves are
$\beta_e(T)=\beta_0\,[1+k_\beta(T-T_0)]$ with $\beta_0=1.5$\,GPa, $T_0=20$,
$k_\beta=-0.005\,$\textdegree C$^{-1}$ (clamped at $0.3$\,GPa);
$\mu(T)=\mu_{40}\exp\!\big(B(1/T_K-1/T_{K,40})\big)$ (Vogel-like) with
$\mu_{40}=41$\,mPa\,s and $B=3100$\,K; and
$C_{\text{leak}}(T)=C_0\,\mu_{40}/\mu(T)$ with $C_0=2\times10^{-13}$\,m$^3$/(s\,Pa).

\begin{table}[h]
\caption{Complete parameter set}
\label{tab:fullparams}
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Symbol & Value & Meaning\\
\midrule
$A_1$ & $7.854\times10^{-3}$\,m$^2$ & cap-side area\\
$A_2$ & $5.391\times10^{-3}$\,m$^2$ & rod-side area\\
$s_\ell$ & $1.0$\,m & stroke\\
$V_{01},V_{02}$ & $2\times10^{-4}$\,m$^3$ & dead volumes\\
$m$ & $200$\,kg & moving mass\\
$P_s$ & $210$\,bar & supply (relief-set)\\
$F_{\text{FS}}$ & $164.9$\,kN & force full scale\\
$\beta_0$ & $1.5$\,GPa & bulk modulus @ $20\,$\textdegree C\\
$\mu_{40}$ & $41$\,mPa\,s & viscosity @ $40\,$\textdegree C\\
$C_0$ & $2\times10^{-13}$ & leak coeff @ $40\,$\textdegree C\\
$C_q$ & $0.67$ & valve discharge coeff\\
ref. freq / amp & $0.04$\,Hz / $1.1\,s_\ell$ & sine reference\\
$K_p,K_i$ & $0.03,0.004$ & PI gains (spool m / m)\\
sample rate & $1$\,kHz & sensor + observer\\
duration & $32$\,s & per run\\
\bottomrule
\end{tabular}
\end{table}

\section{Observer Algorithm}\label{app:obs}
The observer runs at the sample rate. Per step:
\begin{enumerate}
\item Low-pass the measured pressures (40\,Hz) and flows (25\,Hz); form
$\dot{P}_i$ by finite difference of the filtered pressure.
\item Evaluate $\beta_e$, $C_{\text{leak}}$ from the noisy temperature reading.
\item Compute per-chamber velocities $v_1,v_2$ from the rearranged continuity
relations; blend by flow magnitude with a small regulariser so $v\to 0$ when both
flows vanish.
\item Integrate $x \mathrel{+}= v\,\Delta t$, clamped to $[{-}0.02,\,s_\ell{+}0.02]$.
\item If seated (filtered flows small and $\max(P_1,P_2)>0.7\,P_s$) for $>0.3$\,s:
pin $x$ to the corresponding stop; if additionally $\dot P\approx0$ (steady),
update $b_1=Q_1-q_{\text{leak}}$, $b_2=Q_2+q_{\text{leak}}$.
\item Output $\hat{x}$ and $\hat{F}=P_1 A_1-P_2 A_2$.
\end{enumerate}
The steady-pressure gate in step~5 is the single most important detail: without
it, transient seatings (arrive-and-immediately-leave) capture compressibility
flow as if it were bias and corrupt the calibration.

\section{State-Space Model and the Extended Kalman Filter}\label{app:ekf}
This section gives the principled estimator the observer of Sec.~\ref{sec:observer}
approximates: a nonlinear state-space model and the Extended Kalman Filter (EKF)
that is its textbook recursive estimator. We then explain why the deployed
estimator is the reduced complementary observer rather than the full EKF.

\subsection{Nonlinear state-space model}
Take the state, augmented with the two flow-meter biases and the external load so
they are estimated rather than assumed,
\begin{equation}
\boldsymbol{\xi} = \big[\,x,\ v,\ P_1,\ P_2,\ b_1,\ b_2,\ F_L\,\big]^{\!\top}.
\end{equation}
With measured flows $Q_1,Q_2$ and temperature as inputs, the continuous dynamics
$\dot{\boldsymbol\xi}=\mathbf f(\boldsymbol\xi,\mathbf u)$ are
\begin{align}
\dot x   &= v, \\
\dot v   &= \big(P_1 A_1 - P_2 A_2 - F_{\text{fric}}(v) - F_L\big)/m, \\
\dot P_1 &= \tfrac{\beta_e}{V_1(x)}\big[(Q_1-b_1) - A_1 v - q_{\text{leak}}\big], \\
\dot P_2 &= \tfrac{\beta_e}{V_2(x)}\big[(Q_2-b_2) + A_2 v + q_{\text{leak}}\big], \\
\dot b_1 &= \dot b_2 = \dot F_L = 0 \quad(\text{random-walk states}),
\end{align}
with $V_1(x)=V_{01}+A_1x$, $V_2(x)=V_{02}+A_2(s_\ell-x)$ and
$q_{\text{leak}}=C_{\text{leak}}(P_1-P_2)$. Only the two chamber pressures are
measured,
\begin{equation}
\mathbf z = \mathbf h(\boldsymbol\xi) + \boldsymbol\nu
          = [\,P_1,\ P_2\,]^{\!\top} + \boldsymbol\nu .
\end{equation}
Position is observable because $x$ enters the pressure dynamics through the
volumes $V_i(x)$: $\partial \dot P_i/\partial x \neq 0$ whenever there is net
chamber flow, so a pressure innovation carries information about $x$.

\subsection{EKF recursion}
Discretising to $\boldsymbol\xi_{k+1}=\mathbf f_d(\boldsymbol\xi_k,\mathbf u_k)+\mathbf w_k$
with $\mathbf w_k\!\sim\!\mathcal N(0,\mathbf Q)$ and $\mathbf v_k\!\sim\!\mathcal N(0,\mathbf R)$,
the EKF alternates
\begin{align}
\text{predict:}\quad & \hat{\boldsymbol\xi}^- = \mathbf f_d(\hat{\boldsymbol\xi},\mathbf u), \quad
\mathbf P^- = \mathbf F\mathbf P\mathbf F^{\!\top} + \mathbf Q, \\
\text{update:}\quad & \mathbf y = \mathbf z - \mathbf h(\hat{\boldsymbol\xi}^-), \quad
\mathbf S = \mathbf H\mathbf P^-\mathbf H^{\!\top} + \mathbf R, \\
& \mathbf K = \mathbf P^-\mathbf H^{\!\top}\mathbf S^{-1}, \quad
\hat{\boldsymbol\xi} = \hat{\boldsymbol\xi}^- + \mathbf K\mathbf y, \\
& \mathbf P = (\mathbf I-\mathbf K\mathbf H)\mathbf P^-(\mathbf I-\mathbf K\mathbf H)^{\!\top}
            + \mathbf K\mathbf R\mathbf K^{\!\top},
\end{align}
the last in Joseph form for numerical stability. Here
$\mathbf F=\partial\mathbf f_d/\partial\boldsymbol\xi|_{\hat{\boldsymbol\xi}}$ and
$\mathbf H=\partial\mathbf h/\partial\boldsymbol\xi=[\,\mathbf 0\ \mathbf I_2\ \mathbf 0\,]$
(it selects $P_1,P_2$). The load-bearing Jacobian entries are the pressure
sensitivities
\begin{equation}
\frac{\partial \dot P_i}{\partial v} = \mp\frac{\beta_e A_i}{V_i(x)}, \qquad
\frac{\partial \dot P_i}{\partial x} = -\frac{\beta_e A_i}{V_i(x)^2}\,(\,\cdots),
\end{equation}
which couple the pressure measurement to velocity and position, and
$\partial\dot P_i/\partial b_i=-\beta_e/V_i$, which lets the filter learn the
flow-meter biases. $\mathbf Q$ and $\mathbf R$ are set from the process and
sensor-noise magnitudes of Table~\ref{tab:fullparams}.

\subsection{Implementation and conditioning}
A direct EKF on this plant is numerically delicate. The hydraulic stiffness makes
$\beta_e/V_i \sim 10^{12}$--$10^{13}$, so the pressure Jacobian entries above are
enormous, while the state magnitudes span $\sim\!10^{7}$\,(Pa) down to
$\sim\!10^{-2}$\,(m/s) and $\sim\!10^{-5}$\,(m$^3$/s, bias): the covariance
$\mathbf P$ is severely ill-conditioned, and an early prototype that inferred
velocity by integrating the force balance \eqref{eq:newton} diverged. We made a
runnable EKF with four ingredients: (i) \emph{non-dimensionalisation} -- each
state is divided by a physical scale (stroke, $0.3$\,m/s, $P_s$, a bias scale,
$F_{\text{FS}}$) so the scaled covariance is $O(1)$ and Jacobians are taken
numerically in scaled space; (ii) \emph{semi-implicit (symplectic) Euler}
sub-steps for the prediction, stable for the v--P oscillator where forward Euler
at $1$\,kHz is not; (iii) the two \emph{flows as measurements}
($h_Q=\pm A_i v + b_i + q_{\text{leak}} + (V_i/\beta_e)\dot P_i$, the
compressibility term taken from the filtered measured pressure) to make velocity
and bias observable; and (iv) seated \emph{pseudo-measurements} ($x{=}$stop,
$v{=}0$) as the filter's homing. The complementary observer of
Sec.~\ref{sec:observer} can then be read as a structurally-reduced form of this
EKF: its velocity blend is the flow-measurement update on $v$, and its gated
end-stop calibration is the $v{=}0$ pseudo-measurement that separates bias.
Sec.~\ref{sec:ekf} benchmarks the two; the EKF is stable and competitive on force
but worse and less robust on position, so the reduced observer is deployed and
the EKF is kept for the covariance estimate it provides.

\section{Hopsan Integration}\label{app:hopsan}
\subsection{Parameterization}
For each run the model parameters (geometry, the temperature-dependent
$\beta_e$, $C_{\text{leak}}$ and viscous friction, the PI gains and the sine
reference) are overridden, the model is simulated, and the pressure, flow and
position time series are exported to the estimation pipeline.

\subsection{Chamber pressure/flow logging}
In Hopsan's TLM, two ports sharing a node are logged on one side, so the cap and
rod chamber pressures and flows are read on the valve side (PA$\leftrightarrow$cap
$A_1$, PB$\leftrightarrow$rod $A_2$): pressure is identical either side of a node,
and the valve-port flow already equals flow into the chamber. Position is the
shared cylinder/mass position.

\subsection{Controller scaling}
The valve \texttt{in} port is spool position in metres ($|x_v|\le x_{v\max}\approx
0.01$), so $K_p\approx0.03$ maps a ${\sim}1$\,m position error to a ${\sim}0.01$\,m
spool command. An early misconfiguration with $K_p=5$ slammed the valve fully
open, producing an unphysical $2000$\,L/min and $358$\,bar; reducing the gain to
the example's scale fixed it.

\section{Per-Temperature Results}\label{app:results}
Table~\ref{tab:pertemp} gives worst-case (over eight seeds) and mean-RMS errors at
each temperature. Both metrics are essentially flat, confirming that the
end-stop homing reset neutralises accumulated drift across the envelope.

\begin{table}[h]
\caption{Worst-case / mean-RMS error by temperature (8 seeds each)}
\label{tab:pertemp}
\centering
\small
\begin{tabular}{@{}ccccc@{}}
\toprule
$T$ (\textdegree C) & pos max & pos RMS & force 99.5 & force RMS\\
 & (cm) & (cm) & (\% FS) & (\% FS)\\
\midrule
0  & 2.76 & 0.54 & 4.20 & 1.58\\
10 & 2.77 & 0.54 & 4.20 & 1.58\\
20 & 2.79 & 0.54 & 4.20 & 1.58\\
30 & 2.81 & 0.55 & 4.20 & 1.58\\
40 & 2.83 & 0.55 & 4.20 & 1.57\\
50 & 2.85 & 0.55 & 4.20 & 1.57\\
\bottomrule
\end{tabular}
\end{table}

\section{Failure Modes Encountered}\label{app:bugs}
The spike surfaced several issues worth logging, since they shaped the design:
\begin{itemize}
\item \textbf{Position drift (13\,cm).} Long mid-stroke excursions with unlucky
bias draws drifted far past target. Root cause: bias calibration was firing on
transient seats. Fix: steady-pressure gate plus a duty cycle that dwells at
stops.
\item \textbf{Sign of valve-port flow.} The chamber-inflow convention required no
sign flip on \texttt{PA/PB\#Flow}; an initial flip inverted the velocity (negative
correlation with $\dot x$), caught by a correlation check.
\item \textbf{Controller gain.} See \ref{app:hopsan}; a unit mismatch drove
nonphysical flows until corrected.
\item \textbf{(Earlier pure-Python plant, since replaced by Hopsan).} An
explicit-RK4 prototype suffered end-stop-damping instability and cavitation
blow-up, and an EKF that derived velocity by double-integrating the force balance
diverged to over a metre of error. These motivated switching to a TLM solver for
the truth model and to a flow-continuity observer. They are recorded here as part
of the exploration history.
\end{itemize}

\section{Future Directions}\label{app:future}
\begin{itemize}
\item Densify the homing-interval / mismatch stress sweep (more seeds and
frequencies) and report a tail metric (95th percentile / max-over-seeds) with
bootstrap confidence intervals, turning the median/IQR guideline into a
tail-reliability design rule.
\item Improve the EKF (Sec.~\ref{sec:ekf}): better-calibrated covariance and
faster pre-homing convergence (e.g.\ a UKF, or modelling the flow-measurement
error during transients) to make its confidence band trustworthy.
\item Sweep sensor grades to find the cheapest transducers that still pass.
\item Bridge to hardware-style co-simulation via Hopsan's FMU export.
\item Add within-stroke thermal transients with a custom temperature-aware
cylinder component.
\end{itemize}

\section{Session Flow: Human vs.\ AI Effort}\label{app:session}
Per the research-process requirement to preserve what was human input and what
was AI effort, the spike proceeded through these human prompts (paraphrased), each
answered by AI analysis, code and verification:
\begin{enumerate}
\item \emph{``A double-acting cylinder is coupled because flow and pressure at
both ends set position and force, correct?''} AI: clarified the coupling
(flow$\to$velocity$\to$position; pressure$\to$force) and offered a state-space
model.
\item \emph{``Set up a parametric offline simulation showing position error
$\le\pm5$\,cm and force $\le\pm12\,\%$ FS across full stroke and
$0$--$50\,$\textdegree C with realistic noise.''} AI: built a first plant,
sensors, observer and sweep; iterated through stability and drift bugs to a
passing result.
\item \emph{``Can we use Hopsan for the modelling and simulation?''} AI:
researched Hopsan's component library, TLM solver, temperature handling and
interfaces; recommended Hopsan as the plant with the estimator kept in Python.
\item \emph{``Refactor to only use Hopsan as the plant.''} AI: vendored the
Position-Servo example, swapped the reference to a sine source, set up the
parameter overrides, and validated the integration.
\item \emph{``Run the sweep on the real simulator.''} AI: debugged the
chamber pressure/flow logging and controller scaling, and ran the full sweep.
\item \emph{``Treat this chat as a research spike and write it up in IEEE format,
with figures of the sim, Hopsan, and plots.''} AI: built (from source) and
captured the Hopsan model canvas (Fig.~\ref{fig:gui}), generated the figures and
this paper, and ran an external (Codex) review.
\item \emph{``Make the simulation more realistic.''} AI: added per-run plant/
estimator parameter mismatch, Coulomb/stiction friction, finite valve bandwidth
and full transducer dynamics (bandwidth, delay, drift, saturation); added the
ablation study and the homing-interval stress test, which together identified the
homing reset as the load-bearing mechanism.
\item \emph{``Add the EKF theory, then implement a well-scaled EKF and benchmark
it.''} AI: derived the state-space model and EKF recursion (App.~\ref{app:ekf}),
implemented a non-dimensionalised, semi-implicit filter, and benchmarked it
head-to-head against the observer (Sec.~\ref{sec:ekf}).
\item \emph{``Compare with the Simscape lab model; is TLM any different?''} AI:
contrasted the acausal-DAE and transmission-line approaches and added the
solver-rationale discussion (Sec.~\ref{sec:tlm}).
\item \emph{``Make an open-loop variant and extend the sweep to bench scale.''}
AI: built the open-loop model variant, found and fixed an open-loop-only
stop-identification failure (direction-of-travel homing), and validated the bench
case (Sec.~\ref{sec:bench}).
\item \emph{``Validate against the Zhou et al.\ experiment.''} AI: reconfigured
the plant to their physical rig and showed agreement above $15$\,MPa plus
independent confirmation of the low-pressure bulk-modulus mechanism
(Sec.~\ref{sec:zhou}).
\item \emph{``Add hose line losses with manifold-located sensors; sweep hose
length.''} AI: added the volume$+$orifice line model, moved ground-truth force to
the cylinder side, and produced the hose-length design rule (Sec.~\ref{sec:hose}).
\item \emph{``Make it readable from first principles and more visual.''} AI:
added the principle, observer and manifold diagrams (Figs.~\ref{fig:principle},
\ref{fig:obsblock}, \ref{fig:manifold}), a plain-language primer and roadmap, and
tightened the abstract. Final sweep: worst position $2.58$\,cm, operational force
$5.08\,\%$ FS.
\end{enumerate}
All acceptance targets, the choice of Hopsan, and the scope were set by the human;
the modelling, coding, debugging, figure generation and drafting were AI effort
under that direction.

\section{Code and Data Availability}\label{app:code}
The complete harness is open-source at
\begin{center}
\url{https://github.com/rayjyotir5/sensorless-cylinder-hopsan}
\end{center}
comprising the estimator and plant driver
(\texttt{src/cylinder\_sim/}: parameters, Hopsan driver, sensor model, the
complementary observer and the EKF, sweep and plots), the Hopsan models
(\texttt{src/hopsan/sensorless\_cylinder\_plant.hmf} and the open-loop variant)
with a reproducible head-less HopsanCLI build script, the integration tests, and
one figure-generation script per figure in this paper. The referenced
experimental data of \cite{zhou2022pressure} and any third-party lab models are
not redistributed.
